---
title: "Final Project Activity"
author: "Kabin, Nishan, Udita"
date: today
format:
  html:
    theme: flatly
    self-contained: true
    embed-resources: true
    toc: true
execute:
  eval: true  # Evaluate code
  echo: true  # Show code
  warning: false
  message: false
---



```{r}
#| echo: false

# Load packages necessary for this activity using the code below.
library(tidyverse)
library(knitr)
library(skimr)
library(calendR)
library(flextable)
library(naniar)
library(purrr)
library(sf)
```


Importing the data from the csv files; there are four data files, two files containing the data and two files containing the data description .

```{r}
sites_main <- read_csv("pfas_data/pfas_sites.csv")
data_dict_sites <- read_csv("pfas_data/data_dict_sites.csv")
samples_main <- read_csv("pfas_data/pfas_hazard_index.csv")
data_dict_samples <- read_csv("pfas_data/data_dict_hazard.csv")
```

```{r}
data_type_sites <- tibble(Variable = names(sites_main), Type = map_chr(sites_main, typeof))

data_dict_sites <- left_join(data_dict_sites, data_type_sites, by = c("Variable"))
rm(data_type_sites)

data_dict_sites |>
  select(Variable, Type, Description) |>
  flextable() |>
  autofit()
```

```{r}
data_type_samples <- tibble(Variable = names(samples_main), Type = map_chr(samples_main, typeof))

data_dict_samples <- left_join(data_dict_samples, data_type_samples, by = c("Variable"))
rm(data_type_samples)

data_dict_samples |>
  select(Variable, Type, Description) |>
  flextable() |>
  autofit()
```


```{r}
glimpse(sites_main)
glimpse(samples_main)
```



```{r}
skim(sites_main)
skim(samples_main)
```


```{r}
gg_miss_var(sites_main, show_pct = TRUE)
gg_miss_var(samples_main, show_pct = TRUE)
```
Looking at these plots, it's clear that some variables have too much missing data to be useful for our analysis. We decided to drop `site_background`, `drinking_water_information`, `anticipated_activities`, and `location` from the `pfas_sites` dataset because the gaps in these columns would just get in the way of our work.

```{r}
sites <- sites_main |>
  dplyr::select(-site_background, -drinking_water_information, -anticipated_activities, -location)

samples <- samples_main
```


## Data Transformation and Merging

Merging these datasets turned out to be trickier than expected. We couldn't use geoid directly because it identifies counties, not individual sites or samples. And while longitude and latitude were available, they didn't match exactly between the two tables—probably due to measurement differences or rounding.
So we created a new facility column in the samples data to bridge the gap. Here's what we did: for each sample, we calculated the distance to every site in its county and assigned it to the nearest one. For samples missing a geoid, we found their closest site regardless of county boundaries and used that site's geoid and facility name.

This approach required some assumptions about how samples relate to facilities, but it was necessary to get the two datasets linked together for analysis. Obviously, this introduces some uncertainty into the matching, but it was the best option we had given the data constraints. 

```{r}
sites_sf <- sites |>
  st_as_sf(coords = c("longitude", "latitude"),
           crs = 4326,   # WGS84
           remove = FALSE)

samples_sf <- samples |>
  st_as_sf(coords = c("longitude", "latitude"),
           crs = 4326,
           remove = FALSE)

# EPSG:3591
# NAD83(NSRS2007) / Michigan Oblique Mercator

sites_sf   <- st_transform(sites_sf, 3591)
samples_sf <- st_transform(samples_sf, 3591)
```


```{r}
# Remove samples with missing geoid
samples_sf <- samples_sf |>
  filter(!is.na(geoid))

geo_ids <- unique(samples_sf$geoid)

```


```{r}

# For each geoId, match samples to nearest site in that geoId
populate_nearest_facility <- function(samples, sites) {
  if (nrow(sites) == 0) {
    # No PFAS site in this county: keep NA
    samples$nearest_facility <- NA
    samples$distance_m      <- NA
  } else {
    # Index of nearest site in this county for each sample
    idx <- st_nearest_feature(samples, sites)

    samples$nearest_facility <- sites$facility[idx]

    # Distance to that nearest site (element-wise)
    samples$distance_m <- st_distance(samples, sites[idx, ],
                                        by_element = TRUE) |>
      as.numeric()
  }
  samples
}

samples_with_site <- map(geo_ids, function(g){
    samples_g <- samples_sf |> filter(geoid == g)
    sites_g   <- sites_sf   |> filter(geoid == g)
    
    return(populate_nearest_facility(samples_g, sites_g))
} )

```

```{r}
# Combine it into a single dataframe
samples_with_site <- do.call(rbind, samples_with_site)
samples_with_site 

samples_with_site <- samples_with_site |>
  st_drop_geometry()
```

```{r}
samples_site_df <- inner_join(samples_with_site, sites, by = c("nearest_facility" = "facility", "geoid"))
```


